Recursive Solutions to First-Order Model Counting

### 2023-01-25 doc/paper/paper.tex
Start writing a longer talk

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
% \newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

% TODO: double check how the university names should be formatted and the emails
% TODO: read the formatting instructions and their LaTeX source
% TODO: make sure there are no overflow errors
\author{%
Paulius Dilkas$^1$\and
Vaishak Belle$^2$\\
\affiliations
$^1$National University of Singapore, Singapore\\
$^2$University of Edinburgh, UK\\
\emails
paulius.dilkas@nus.edu.sg,
vbelle@ed.ac.uk
}

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s)~\cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals~\cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances
become solvable in polynomial time~\cite{DBLP:conf/nips/Broeck11}. Lifted
inference algorithms are typically used on probabilistic models such as
probabilistic programming languages
\cite{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17}, Markov logic
networks
\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical~\cite{DBLP:journals/ml/KimmigMG15} and statistical
relational~\cite{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by WFOMC
\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC tends to receive more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory~\cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

% TODO: have ~ in front of every~\cite
% TODO: fix doom emacs warnings
Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the domain(s)
\cite{jaeger2012liftability}. Over more than a decade, many classes of instances
were shown to be liftable
\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two results came out, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and \SRU{}
\cite{DBLP:conf/nips/KazemiKBP16}, \Uone{}~\cite{DBLP:conf/lics/KuusistoL18},
\FOtwo{} with tree axioms~\cite{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the
two-variable fragment with counting quantifiers)
\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On the empirical
front, there are several open-source implementations of exact WFOMC algorithms:
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}, probabilistic theorem
proving~\cite{DBLP:journals/cacm/GogateD16}, and \textsc{L2C}
\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is supported by
\textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving~\cite{DBLP:journals/cacm/GogateD16}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is previous work
 ~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/ijcai/Broeck16} showing how FOMC
  can compute factorials in principle, but an algorithm that can lift such an
  instance in practice does not exist.} We claim that this shortcoming is due to
the inability of these algorithms to construct recursive solutions. The topic of
recursion in the context of WFOMC has been studied before but in limited ways.
Barv{\'{\i}}nek et al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to
generate numerical data that is then used to conjecture recurrence relations
that explain that data. Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11}
introduced the idea of \emph{domain recursion}. Intuitively, domain recursion
partitions a domain of size $n$ into a single explicitly named constant and the
remaining domain of size $n-1$. However, many stringent conditions are enforced
to ensure that the search for a tractable solution always terminates.

@@

Along with many node types defined previously
\cite{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11}, \textsc{Crane}
uses three more:
\begin{itemize}
  \item a type for constraint removal denoted by $\CR$,
  \item a type for generalised domain recursion denoted by $\GDR$ (both with
        out-degree one),
  \item and $\star$---a placeholder type (with out-degree zero) for nodes that
        are going to be replaced.
\end{itemize}
Similarly to Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}, we
write $T_p$ for an FCG that has a node with the label $T_p$ (i.e., type $T$ and
parameter(s) $p$) and $\star$'s as all of its direct successors. See
\cref{fig:examplefcg} for an example FCG\@. Its source node has out-degree 1,
label $\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$.

@@

A \emph{(compilation) rule} takes a formula and returns a set of chips. The
cardinality of this set is the number of different ways in which the rule can be
applied to the input formula. While \textsc{ForcLift}
\cite{DBLP:conf/ijcai/BroeckTMDR11} heuristically chooses one of them, in an
attempt to not miss a solution, \textsc{Crane} returns them all. In particular,
if a rule returns an empty set, then that rule does not apply to the formula.

@@

@@

\section{Experimental Results}\label{sec:results}
% TODO: the thesis has an improved version of this section
% TODO: when describing experimental results, say that a solution for the last
% (unsolved) instance can be found, but is too complicated to evaluate by hand.
% Add that the linear order axiom is also liftable. Cite the same paper in the
% introduction as well.

\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!10}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!10}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!10}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!10}\cmark & \xmark & & $\min\{\, m, n \,\}^2$ & --- & $mn$ \\
    \rowcolor{gray!10}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions.
    Here, $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that the algorithm was not able to find a solution.}\label{tbl:results}
\end{table*}
% TODO: It is a standard to discuss a few lifted theories like the smokers and
% friends example. Can you also include some empirical results on such theories
% as well to appeal to the sanity of the reader that it actually works for
% problems in frameworks such as Markov log networks

We compare \textsc{Crane} and \textsc{ForcLift}
\cite{DBLP:conf/ijcai/BroeckTMDR11} on their ability to count various kinds of
functions. This class of instances is chosen because of its simplicity and
ubiquity and the inability of state-of-the-art WFOMC algorithms to solve many
such instances. Note that other WFOMC algorithms---\textsc{L2C}
\cite{DBLP:conf/kr/KazemiP16} and probabilistic theorem proving
\cite{DBLP:journals/cacm/GogateD16}---are unable to solve any of the instances
that \textsc{ForcLift} fails on. We begin by describing how such
function-counting problems can be expressed in FOL\@. \textsc{ForcLift} then
translates these sentences in FOL to formulas as defined in \cref{def:formula}.

% Describe what kind of instances we're investigating
Let $p \in \Gamma \times \Delta$ be a predicate. To restrict all relations
representable by $p$ to just functions from $\Gamma$ to $\Delta$, in FOL one
might write
\[
  \forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }p(X, Y) \land p(X, Z) \implies Y = Z
\]
and
\begin{equation}\label{eq:def2}
  \forall X \in \Gamma\text{. }\exists Y \in \Delta\text{. }p(X, Y).
\end{equation}
The former sentence says that one element of $\Gamma$ can map to at \emph{most}
one element of $\Delta$, and the latter sentence says that each element of
$\Gamma$ must map to at \emph{least} one element of $\Delta$. One can then add
\[
  \forall X,Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }p(X, Y) \land p(Z, Y) \implies X = Z
\]
to restrict $p$ to injections or
\[
  \forall Y \in \Delta\text{. }\exists X \in \Gamma\text{. }p(X, Y)
\]
to ensure surjectivity or remove \cref{eq:def2} to consider partial functions.
Lastly, one can replace all occurrences of $\Delta$ with $\Gamma$ to model
endofunctions (i.e., functions with the same domain and codomain) instead.

In our experiments, we consider all sixteen combinations of these properties,
i.e., injectivity, surjectivity, partiality, and endo-. \textsc{ForcLift} is
always run until it terminates. \textsc{Crane} is run until either five
solutions are found or the search tree reaches height 6 (which happens in at
most a few seconds). If successful, \textsc{ForcLift} generates a circuit, and
\textsc{Crane} generates one or more (complete) FCGs. In both cases, we manually
convert the resulting graphs into definitions of functions as described in
\cref{sec:interpret}. We then assess the complexity of each solution and pick
the best if \textsc{Crane} returns several solutions of varying complexities.
When assessing the complexity of each such definition, we make two assumptions.
First, we can compute the binomial coefficient $\binom{n}{k}$ in $\Theta(nk)$
time. Second, techniques such as dynamic programming and memoization are used to
avoid recomputing the same binomial coefficient or function call multiple times.

The experimental results are in \cref{tbl:results}. Previous work often compares
WFOMC algorithms by running them on a few instances with increasing domain sizes
and measuring runtime
\cite{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11,DBLP:conf/aaai/BroeckD12}.
However, we can do better than that and identify the exact worst-case asymptotic
complexity of a solution produced by either \textsc{Crane} or \textsc{ForcLift}.
The best-known asymptotic complexity for computing total surjections is by
Earnest~\shortcite{30049}. All other best-known complexity results are inferred
from the formulas and programs on the on-line encyclopedia of integer sequences
\cite{oeis}. On instances that could already be solved by \textsc{ForcLift}, the
two algorithms perform equally well. However, \textsc{Crane} can also solve all
but one instances that \textsc{ForcLift} fails on in at most cubic time.

Let us examine the case of counting partial (non-endomorphic) injections more
closely. The FCG in \cref{fig:examplefcg,example:interpretation} counts partial
injections and is responsible for the $mn$ entry in the table. For a complete
solution, \cref{eq:solution} must be combined with the base case $f(0, n) = 1$
for all $n \in \mathbb{N}_{0}$. This base case says that the empty partial map
is the only partial injection with an empty domain. Finally, note that $f(m, n)$
can be evaluated in $\Theta(mn)$ time by a dynamic programming algorithm that
computes $f(i, j)$ for all $i = 0, \dots, m$ and $j = 0, \dots, n$.

@@

Recall that \Ctwo{} is the class of formulas with counting quantifiers and at
most two variables. \Ctwo{} was recently shown to be liftable
\cite{DBLP:journals/jair/Kuzelka21} but without providing a usable (W)FOMC
algorithm. Since the tasks of counting injections and bijections fall into the
\Ctwo{} fragment, we can conjecture the following.

@@

However, the most important direction for future work is to fully automate this
new process. First, we need an algorithm that transforms FCGs into definitions
of functions. Formalising this process would also allow us to prove the
correctness of the new compilation rules in constructing FCGs that indeed
compute the right (weighted) model counts. Second, these definitions must be
simplified before they can be used, perhaps by a computer algebra system. Third,
we need a way to find the base cases for the recursive definitions provided by
\textsc{Crane}. Fourth, since the first solution found by \textsc{Crane} is not
always optimal in terms of its complexity, an automated way to determine the
asymptotic complexity of a solution would be helpful as well. Achieving these
goals would enable \textsc{Crane} to automatically constructing efficient ways
to compute various functions and sequences. In addition to the potential impact
on areas of artificial intelligence (AI) such as statistical relational AI
\cite{DBLP:series/synthesis/2016Raedt}, \textsc{Crane} could be beneficial to
research in combinatorics as well~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

\appendix
\section{Solutions Found by \textsc{Crane}}

@@

% TODO: in the acknowledgements, mention that most of the work was done at UoE

>>>>

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
% \newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

% TODO: double check how the university names should be formatted and the emails
% TODO: read the formatting instructions and their LaTeX source
% TODO: make sure there are no overflow errors
% TODO: for arbitrary instances of '2022' (e.g., in citations), make it '2023'
\author{%
Paulius Dilkas$^1$\and
Vaishak Belle$^2$\\
\affiliations
$^1$National University of Singapore, Singapore\\
$^2$University of Edinburgh, UK\\
\emails
paulius.dilkas@nus.edu.sg,
vbelle@ed.ac.uk
}

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s)~\cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals~\cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances become
solvable in polynomial time~\cite{DBLP:conf/nips/Broeck11}. Lifted inference
algorithms are typically used on probabilistic models such as probabilistic
programming
languages~\cite{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17},
Markov logic
networks~\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical~\cite{DBLP:journals/ml/KimmigMG15} and statistical
relational~\cite{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by
WFOMC~\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC tends to receive more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory~\cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

% TODO: fix doom emacs warnings
Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the
domain(s)~\cite{jaeger2012liftability}. Over more than a decade, many classes of
instances were shown to be
liftable~\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two results came out, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and
\SRU{}~\cite{DBLP:conf/nips/KazemiKBP16},
\Uone{}~\cite{DBLP:conf/lics/KuusistoL18}, \FOtwo{} with tree
axioms~\cite{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the two-variable
fragment with counting
quantifiers)~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On
the empirical front, there are several open-source implementations of exact
WFOMC algorithms: \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11},
probabilistic theorem proving~\cite{DBLP:journals/cacm/GogateD16}, and
\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is supported by
\textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving~\cite{DBLP:journals/cacm/GogateD16}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is previous
  work~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/ijcai/Broeck16} showing how
  FOMC can compute factorials in principle, but an algorithm that can lift such
  an instance in practice does not exist.} We claim that this shortcoming is due
to the inability of these algorithms to construct recursive solutions. The topic
of recursion in the context of WFOMC has been studied before but in limited
ways. Barv{\'{\i}}nek et al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC
to generate numerical data that is then used to conjecture recurrence relations
that explain that data. Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11}
introduced the idea of \emph{domain recursion}. Intuitively, domain recursion
partitions a domain of size $n$ into a single explicitly named constant and the
remaining domain of size $n-1$. However, many stringent conditions are enforced
to ensure that the search for a tractable solution always terminates.

@@

Along with many node types defined
previously~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11},
\textsc{Crane} uses three more:
\begin{itemize}
  \item a type for constraint removal denoted by $\CR$,
  \item a type for generalised domain recursion denoted by $\GDR$ (both with
        out-degree one),
  \item and $\star$---a placeholder type (with out-degree zero) for nodes that
        are going to be replaced.
\end{itemize}
Similarly to Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}, we
write $T_p$ for an FCG that has a node with the label $T_p$ (i.e., type $T$ and
parameter(s) $p$) and $\star$'s as all of its direct successors. See
\cref{fig:examplefcg} for an example FCG\@. Its source node has out-degree 1,
label $\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$.

@@

A \emph{(compilation) rule} takes a formula and returns a set of chips. The
cardinality of this set is the number of different ways in which the rule can be
applied to the input formula. While
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} heuristically chooses one
of them, in an attempt to not miss a solution, \textsc{Crane} returns them all.
In particular, if a rule returns an empty set, then that rule does not apply to
the formula.

@@

TODO: connecting sentence?

\begin{proposition}[Correctness of GDR]
  Let $\phi$ be the formula used as input to \cref{alg:domainrecursion} and
  $\phi'$ be one of the formulas constructed by the algorithm. Then
  $\phi \equiv \phi'$.
\end{proposition}
\begin{proof}
  TODO
  Assume that \cref{line:condition} selects domain $\Omega \in \mathcal{D}$.
  \[
    \forall X \in \Omega\text{. } \psi \equiv \psi[x/\{\, X \,\}] \land \forall X \in \Omega\text{. } X \ne x \implies \psi
  \]
\end{proof}

@@

\section{Empirical Results}\label{sec:results}
% TODO: when describing experimental results, say that a solution for the last
% (unsolved) instance can be found, but is too complicated to evaluate by hand.
% Add that the linear order axiom is also liftable. Cite the same paper in the
% introduction as well.

% TODO: make sure this table floats to the right page and look at other floating
% objects too (e.g., is [t] on?)
\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!25}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!25}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!25}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!25}\cmark & \xmark & & ${\min\{\, m, n \,\}}^2$ & --- & $mn$ \\
    \rowcolor{gray!25}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions,
    where $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that the algorithm was not able to find a solution.}\label{tbl:results}
\end{table*}
% TODO: It is a standard to discuss a few lifted theories like the smokers and
% friends example. Can you also include some empirical results on such theories
% as well to appeal to the sanity of the reader that it actually works for
% problems in frameworks such as Markov log networks

We compare \textsc{Crane} and
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} on their ability to count
various kinds of functions. This class of instances is chosen because of its
simplicity and ubiquity and the inability of state-of-the-art WFOMC algorithms
to solve many such instances. Note that other WFOMC
algorithms---\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16} and probabilistic
theorem proving~\cite{DBLP:journals/cacm/GogateD16}---are unable to solve any of
the instances that \textsc{ForcLift} fails on. We begin by describing how such
function-counting problems can be expressed in FOL\@. \textsc{ForcLift} then
translates these sentences in FOL to formulas as defined in \cref{def:formula}.

% Describe what kind of instances we're investigating
Let $p \in \Gamma \times \Delta$ be a predicate. To restrict all relations
representable by $p$ to just functions from $\Gamma$ to $\Delta$, in FOL one
might write
\[
  \forall X \in \Gamma\text{. }\forall Y,Z \in \Delta\text{. }p(X, Y) \land p(X, Z) \implies Y = Z
\]
and
\begin{equation}\label{eq:def2}
  \forall X \in \Gamma\text{. }\exists Y \in \Delta\text{. }p(X, Y).
\end{equation}
The former sentence says that one element of $\Gamma$ can map to at \emph{most}
one element of $\Delta$, and the latter sentence says that each element of
$\Gamma$ must map to at \emph{least} one element of $\Delta$. One can then add
\[
  \forall X,Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }p(X, Y) \land p(Z, Y) \implies X = Z
\]
to restrict $p$ to injections or
\[
  \forall Y \in \Delta\text{. }\exists X \in \Gamma\text{. }p(X, Y)
\]
to ensure surjectivity or remove \cref{eq:def2} to consider partial functions.
Lastly, one can replace all occurrences of $\Delta$ with $\Gamma$ to model
endofunctions (i.e., functions with the same domain and codomain) instead.

% the process: both for running algorithms and for determining complexities
In our experiments, we consider all sixteen combinations of these properties,
i.e., injectivity, surjectivity, partiality, and endo-. We run each algorithm
once on each of the sixteen instances.\footnote{While domain sizes have to be
  specified as part of the input, they are immaterial to the search procedure.}
\textsc{ForcLift} is always run until it terminates; \textsc{Crane} is run until
either five solutions are found or the search tree reaches height
six.\footnote{The search tree has a high branching factor, so exploring all
  nodes at depth five takes at most a few seconds, whereas doing the same for
  depth six can be computationally infeasible in some cases.} If successful,
\textsc{ForcLift} generates a circuit, and \textsc{Crane} generates one or more
(complete) FCGs. In both cases, we manually convert the resulting graphs into
definitions of functions as described in \cref{sec:interpret}. We then assess
the complexity of each solution by hand and pick the best in case \textsc{Crane}
returns several solutions of varying complexities.

% how the complexity of a solution is determined
Previous work often compares WFOMC algorithms by running them on a few instances
with increasing domain sizes and measuring
runtime~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11,DBLP:conf/aaai/BroeckD12}.
However, we can do much better and identify the exact worst-case asymptotic
complexity of a solution produced by either \textsc{Crane} or \textsc{ForcLift}.
The asymptotic complexity of a solution can be determined by considering the
total number of arithmetic operations needed to follow the definitions of
constituent functions without recomputing the same quantity multiple times. In
particular, we assume that each function call and binomial coefficient is
computed at most once. We also assume that computing $\binom{n}{k}$ takes
$\Theta(nk)$ time. For example, the complexity of \cref{eq:solution} is
$\Theta(mn)$ since $f(m, n)$ can be computed by a dynamic programming algorithm
that computes $f(i, j)$ for all $i = 0, \dots, m$ and $j = 0, \dots, n$, taking
a constant amount of time on the computation of each $f(i, j)$.

These results are summarised in \cref{tbl:results}, where we compare the
solutions found by both algorithms to the best known ways of computing the same
quantities. The best-known asymptotic complexity for computing total surjections
is by Earnest~\shortcite{30049}. All other best-known complexity results are
inferred from the formulas and programs on the on-line encyclopedia of integer
sequences~\cite{oeis}. On instances that could already be solved by
\textsc{ForcLift}, the two algorithms perform equally well. However,
\textsc{Crane} can also solve all but one instances that \textsc{ForcLift} fails
on in at most cubic time. See \cref{app:solutions} for the exact solutions
produced by \textsc{Crane} that correspond to the complexity results in
\cref{tbl:results}.

@@

Recall that \Ctwo{} is the class of formulas with counting quantifiers and at
most two variables. \Ctwo{} was recently shown to be
liftable~\cite{DBLP:journals/jair/Kuzelka21} but without providing a usable
(W)FOMC algorithm. Since the tasks of counting injections and bijections fall
into the \Ctwo{} fragment, we can conjecture the following.

@@

However, the most important direction for future work is to fully automate this
new process. First, we need an algorithm that transforms FCGs into definitions
of functions. Formalising this process would also allow us to prove the
correctness of the new compilation rules in constructing FCGs that indeed
compute the right (weighted) model counts. Second, these definitions must be
simplified before they can be used, perhaps by a computer algebra system. Third,
we need a way to find the base cases for the recursive definitions provided by
\textsc{Crane}. Fourth, since the first solution found by \textsc{Crane} is not
always optimal in terms of its complexity, an automated way to determine the
asymptotic complexity of a solution would be helpful as well. Achieving these
goals would enable \textsc{Crane} to automatically constructing efficient ways
to compute various functions and sequences. In addition to the potential impact
on areas of artificial intelligence (AI) such as statistical relational
AI~\cite{DBLP:series/synthesis/2016Raedt}, \textsc{Crane} could be beneficial to
research in combinatorics as well~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

\appendix
\section{Solutions Found by \textsc{Crane}}\label{app:solutions}

% TODO: fix the introductory description

@@

% TODO: in the acknowledgements, mention that most of the work was done at UoE
% TODO: should the appendix be in the same file or not?
